Об использовании деревьев решений для выявления областей притяжения локальных минимумов в параллельном алгоритме глобальной оптимизации \footnote{Статья рекомендована к публикации программным комитетом Международной научной конференции <<Параллельные вычислительные технологии (ПаВТ) 2023>>}

### 2023-07-27 2023/vestnik.susu/example.tex
Статья в вестник Южно-Уральского государственного университета

%%%%%%%%%%%%%%%%%%%% author.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% sample root file for your "contribution" to a proceedings volume
%
% Use this file as a template for your own input.
%
%%%%%%%%%%%%%%%% Springer %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass{svproc}
\usepackage{marvosym}
%
% RECOMMENDED %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\usepackage{graphicx}

% to typeset URLs, URIs, and DOIs
\usepackage{url}
\usepackage{hyperref}
\usepackage[english]{babel} 
\usepackage{amsmath} 
\def\UrlFont{\rmfamily}
\providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}

\def\orcidID#1{\unskip$^{[#1]}$}
\def\letter{$^{\textrm{(\Letter)}}$}

\begin{document}
\mainmatter              % start of a contribution
%
\title{On using the decision trees to identify the local extrema in parallel global optimization algorithms}
%
\titlerunning{On using the decision trees to identify the local extrema }  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Konstantin Barkalov\inst{1}\letter\orcidID{0000-0001-5273-2471} \and Ilya Lebedev\inst{1}\letter\orcidID{0000-0002-8736-0652} \and Dmitry Silenko\inst{1}\letter\orcidID{0000-0002-2578-9699}}
%
\authorrunning{Konstantin Barkalov et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{K.A.~Barkalov and I.G.~Lebedev, and D.I.~Silenko}
%
\institute{Lobachevsky State University of Nizhny Novgorod, Nizhny Novgorod, Russia \\
\email{ \{konstantin.barkalov,ilya.lebedev,silenko\}@itmm.unn.ru}
}

\maketitle              % typeset the title of the contribution

\begin{abstract}
In the present work, the solving of the multidimensional global optimization problems using decision tree to reveal the attractor regions of the local minima is considered. The objective function of the  problem is defined as a ``black box''.  We assume it only to satisfy Lipschitz condition with unknown constant.  Global search algorithm is applied for the search of global minimum in the problems of such type. In the present work, we propose a method for selecting the vicinity of local extrema of the objective  function based on analysis of accumulated search information. Conducting such an analysis using machine learning techniques allows making a decision to run a local  method, which can speed up the convergence of the algorithm.  This suggestion was confirmed by the results of numerical experiments demonstrating the speedup  when solving a series of test problems.
% We would like to encourage you to list your keywords within
% the abstract section using the \keywords{...} command.
\keywords{Global optimization $\cdot$ Multiextremal functions $\cdot$ Parallel computing $\cdot$ Machine learning $\cdot$ Decision tree}
\end{abstract}
%

\section{Introduction}

In the present work, we consider the parallel algorithms for solving the multidimensional global optimization problems. Such problems often arise in the cases, when it is necessary to select the values  of parameters of the mathematical model being investigated, in which the simulation results best fit the experimental data. For solving the problems of this class, many algorithms are known: from the metaheuristic algorithms, based on the  idea of random search \cite{Ferreiro2013,Garcia2014,Langdon2011}, to the deterministic algorithms guarantying convergence to the global minimum \cite{Evtushenko2009,He2008,Paulavicius2011}.

Since in real global optimization problems each computing of the function value (hereafter called  \textit{a trial}) is a computation-costly operation, one has to reduce the number of such trials. It  can be achieved by intentional choice of variants in the course of search for the optimal solution cutting  off unpromising search subdomains and investigating only these ones, in which the solution of the  problem can be found. The global search algorithm (GSA) is based on this idea \cite{Strongin2000}. In  the present work, we tried to combine GSA and a local optimization method (Hooke-Jeeves pattern search method \cite{HookJeeves}) to reduce the number of trials executed. The decision on the run of the local method will be made using a decision tree.

The local optimization algorithms are intended for determining only one of local extrema within a set  of feasible solutions, in which the objective function takes the extremal (maximum or minimum) value.  Among the local methods,  the zero-order methods and the gradient methods (of the 1\textsuperscript{st} order and of the  2\textsuperscript{nd} one) are traditionally distinguished \cite{Nocedal}. Note that the application of the first-order methods (not to mention the second one) in the problems  with the ``black-box'' type functions is difficult since the problem of numerical estimating the gradient arises here \cite{Kelley}. Therefore, further we will consider only the zero-order methods.

@@

The substantiation of this method of organization of computations see in \cite{Strongin2000,Barkalov2016}. The  modifications taking into account the presence of inequality constraints as well as the information on the derivative of the objective function are presented in \cite{Barkalov2002,Gergel1997}.

@@

The Hooke--Jeeves pattern search algorithm belongs to the class of zero-order methods. Its computation rules are a  combination of the investigating search (to select the direction) and search in the direction selected  \cite{Himmelblau72}.

@@

\begin{figure}
	\begin{center}
		\begin{minipage}[h]{0.8\linewidth}
			\includegraphics[width=1\linewidth]{figure/fig5.png}
			\caption{An objective function $\varphi(y)$ from the test class} %% подпись к рисунку
			\label{fig:fig2}
		\end{minipage}
	\end{center}
\end{figure}

\begin{figure}
	\begin{center}
		\begin{minipage}[h]{0.8\linewidth}
			\includegraphics[width=1\linewidth]{figure/fig4.png}
			\caption{A piecewise-constant  approximation $\psi(y)$ of the function $\varphi(y)$} %% подпись к рисунку
			\label{fig:fig2_2}
		\end{minipage}
	\end{center}
\end{figure}

When implementing the algorithm for finding the attraction areas of the local  minima, we used the algorithms from OpenCV library to construct decision tree. OpenCV is an open source library of the algorithms of computer vision, image processing, and  general-purpose numerical algorithms. More details on the decision tree can be found  in \cite{Brahmbhatt2013}.

@@

%Ниже приведены результаты сравнения двух параллельных алгоритмов --- асинхронного GSA и его модификации с использованием decision tree и локального метода Хука-Дживса. Численное сравнение проводилось на классах функций Simple и Hard размерности 2, 3, 4 и 5 из \cite{Sergeyev2006}. Критерием остановки служило попадание точки очередного испытания trial в $\varepsilon$ окрестность истинного глобального минимума. 
%В таблице \ref{tab:2} представлено среднее количество итераций и ускорение по времени относительно последовательного запуска. Распараллеливание  осуществлялось с применением  MPI. При этом запуск производился на 4 процессах, а следовательно, на каждой итерации вычислялось 3 точки. Значение в скобках (если оно есть) показывает количество не решенных задач. Это означает, что было достигнут максимум по числу проводимых итераций, но в окрестность глобального минимума точка так и не попала.

Below, the results of comparing two parallel algorithms --- asynchronous global search algorithm  (AGSA) and its modification using the decision tree to find the attractor areas of the local minima  (Decision Tree) are presented. The numerical comparison was performed on two classes of the GKLS  functions (Simple and Hard from \cite{Sergeyev2006}) of dimensionalities 2, 3, 4, and 5. 
These two classes differ in the size of the attraction region of the global minimum point. 
For the simple class of problems, the radius of the attraction region is three times larger than for a complex one.

@@

In addition to using decision trees to identify attraction regions for local extrema of multiextremal functions, we also plan to use machine learning methods to separate the variables of the problem being solved. In many applied optimization problems the dependence of the objective function on some parameters is either linear or unimodal. Separating such variables into a special group and solving the problem using a parallel recursive optimization scheme \cite{Barkalov2020_1} can reduce the time for solving the problem by orders of magnitude compared to using a global search for all variables at once.

\medskip

\textbf{Acknowledgments}. This study was supported by the Russian Science Foundation, project No.\,21-11-00204 .

%
% ---- Bibliography ----
%
\bibliographystyle{spmpsci}
\bibliography{bibliography}{}

>>>>

\documentclass{cmi}
\usepackage{upgreek}

% Данные поля заполняются редакцией журнала
%\makeatletter
%\newcommand{\@volume}{}
%\newcommand{\@issue}{}
%\newcommand{\@num}{}
%\newcommand{\@pages}{}

% Подключение bib файлов
\addbblresources{biblio_ru.bib}{biblio_en.bib}

\begin{document}

% ЗАПОЛНЯЕТСЯ АВТОРАМИ: Релевантные коды УДК статьи
% (см. http://teacode.com/online/udc/)
\classify{УДК  519.853.4}

\renewcommand*{\thefootnote}{\fnsymbol{footnote}} %to make a number in end of title delete this line

\title{Об использовании деревьев решений для выявления областей притяжения локальных минимумов в параллельном алгоритме глобальной оптимизации%
	\footnote{Статья рекомендована к публикации программным комитетом Международной научной конференции <<Параллельные вычислительные технологии (ПаВТ) 2023>>}%
}

\authors{К.А.~Баркалов$^{1}$, И.Г.~Лебедев$^{1}$, Д.И.~Силенко$^{1}$}

\organizations{$^{1}$Нижегородский государственный университет им. Н.И. Лобачевского\\
	(603022, Н.Новгород, проспект Гагарина, ~23)}

\newcommand{\emailsinsert}{ \href{mailto:barkalov@vmk.unn.ru}{barkalov@vmk.unn.ru}, \href{mailto:ilya.lebedev@itmm.unn.ru}{ilya.lebedev@itmm.unn.ru}, \href{mailto:silenko@itmm.unn.ru}{silenko@itmm.unn.ru} }

% ЗАПОЛНЯЕТСЯ АВТОРАМИ: Дата поступления статьи в редакцию 
\newcommand{\rdate}{28.07.2023}
\emails{\emailsinsert}
\redactiondate{\rdate}

\renewcommand*{\thefootnote}{\arabic{footnote}}
\setcounter{footnote}{0}

% Оформление аннотации
\begin{abstract}%
В работе рассматривается решение многомерных задач глобальной оптимизации с использованием деревьев решений для выявления областей притяжения локальных минимумов. Целевая функция представлена как «черный ящик», для которой известно, лишь что она удовлетворяет условию Липшица с неизвестной константой. Для решения этой задачи применятся алгоритм глобального поиска. В работе предлагается метод определения окрестности локальных минимумов целевой функции на основе анализа накопленной поисковой информации. Проведение такого анализа с использованием методов машинного обучения позволяет принять решение о запуске локального метода, что может ускорить сходимость алгоритма. Данный подход был подтвержден результатами численных экспериментов, демонстрирующих ускорение при решении набора тестовых задач.

\keywords{Глобальная оптимизация, Многоэкстремальные функции, Параллельные вычисления, Машинное обучение, Дерево решений}
\end{abstract}

%Образец цитирования статьи
\forcitation{Баркалов К.А., Лебедев  И.Г., Силенко  Д.И. Об использовании деревьев решений для выявления областей притяжения локальных минимумов в параллельном алгоритме глобальной оптимизации}

%Верхние колонтитулы
\markboth{К.А.~Баркалов, И.Г.~Лебедев, Д.И.~Силенко}{Об использовании деревьев решений в параллельном алгоритме глобальной оптимизации \ldots}

\vspace{1em}
{\it Работа выполнена при поддержке Российского научного фонда, проект № 21-11-00204 .}

\insertbiblio

% Сведения об авторах (Ф.И.О.(полностью), степень, звание, кафедра (отдел), вуз, e-mail)
Первый Александр Васильевич, д.ф.-м.н., профессор, кафедра системного программирования, Южно-Уральский государственный университет (национальный исследовательский университет) (Челябинск, Российская Федерация)

Второй Владимир Григорьевич, к.т.н., доцент, кафедра суперкомпьютеров и квантовой информатики, Московский государственный университет имени М.В.~Ломоносова (Москва, Российская Федерация)

%\newpage

{\phantom{eee} \hrule \vskip 4 pt}
\classify{} % Код MSC не нужен
%\classify{MSC 00000} % Коды MSC см. http://www.ams.org/msc/

\title{On using the decision trees to identify the local extrema in parallel global optimization algorithms}

\authorsen{Konstantin Barkalov$^{1}$, Ilya Lebedev$^{1}$,  Dmitry Silenko$^{1}$}

\organizations{$^{1}$Lobachevsky State University of Nizhny Novgorod\\
	(23 Prospekt Gagarina, Nizhny Novgorod, 603022 Russia)}

\emails{\emailsinsert}
\redactiondateen{\rdate}

\begin{abstract}%
In the present work, the solving of the multidimensional global optimization problems using decision tree to reveal the attractor regions of the local minima is considered. The objective function of the  problem is defined as a ``black box''.  We assume it only to satisfy Lipschitz condition with unknown constant.  Global search algorithm is applied for the search of global minimum in the problems of such type. In the present work, we propose a method for selecting the vicinity of local extrema of the objective  function based on analysis of accumulated search information. Conducting such an analysis using machine learning techniques allows making a decision to run a local  method, which can speed up the convergence of the algorithm.  This suggestion was confirmed by the results of numerical experiments demonstrating the speedup  when solving a series of test problems.

\keywordsen{Global optimization, Multiextremal functions, Parallel computing, Machine learning, Decision tree}
\end{abstract}

\forcitationen{Pervyi~A.B., Vtoroi~V.G. The \LaTeXe{} Template for Paper Submission for the ``Computational Mathematics and Software Engineering'' Series}

% Заголовки разделов формируются при помощи команд  \section{}, \subsection{}, \subsubsection{}
\section{Introduction}

In the present work, we consider the parallel algorithms for solving the multidimensional global optimization problems. Such problems often arise in the cases, when it is necessary to select the values  of parameters of the mathematical model being investigated, in which the simulation results best fit the experimental data. For solving the problems of this class, many algorithms are known: from the metaheuristic algorithms, based on the  idea of random search \bblcite{Ferreiro2013,Garcia2014,Langdon2011}, to the deterministic algorithms guarantying convergence to the global minimum \bblcite{Evtushenko2009,He2008,Paulavicius2011}.

Since in real global optimization problems each computing of the function value (hereafter called  \textit{a trial}) is a computation-costly operation, one has to reduce the number of such trials. It  can be achieved by intentional choice of variants in the course of search for the optimal solution cutting  off unpromising search subdomains and investigating only these ones, in which the solution of the  problem can be found. The global search algorithm (GSA) is based on this idea \bblcite{Strongin2000}. In  the present work, we tried to combine GSA and a local optimization method (Hooke-Jeeves pattern search method \bblcite{HookJeeves}) to reduce the number of trials executed. The decision on the run of the local method will be made using a decision tree.

The local optimization algorithms are intended for determining only one of local extrema within a set  of feasible solutions, in which the objective function takes the extremal (maximum or minimum) value.  Among the local methods,  the zero-order methods and the gradient methods (of the 1\textsuperscript{st} order and of the  2\textsuperscript{nd} one) are traditionally distinguished \bblcite{Nocedal}. Note that the application of the first-order methods (not to mention the second one) in the problems  with the ``black-box'' type functions is difficult since the problem of numerical estimating the gradient arises here \bblcite{Kelley}. Therefore, further we will consider only the zero-order methods.

@@

The substantiation of this method of organization of computations see in \bblcite{Strongin2000,Barkalov2016}. The  modifications taking into account the presence of inequality constraints as well as the information on the derivative of the objective function are presented in \bblcite{Barkalov2002,Gergel1997}.

@@

The Hooke--Jeeves pattern search algorithm belongs to the class of zero-order methods. Its computation rules are a  combination of the investigating search (to select the direction) and search in the direction selected  \bblcite{Himmelblau72}.

@@

\begin{figure}
	\begin{center}
		\begin{minipage}[h]{0.7\linewidth}
			\includegraphics[width=1\linewidth]{figure/fig5.png}
			\caption{An objective function $\varphi(y)$ from the test class} %% подпись к рисунку
			\label{fig:fig2}
		\end{minipage}
	\end{center}
\end{figure}

\begin{figure}
	\begin{center}
		\begin{minipage}[h]{0.7\linewidth}
			\includegraphics[width=1\linewidth]{figure/fig4.png}
			\caption{A piecewise-constant  approximation $\psi(y)$ of the function $\varphi(y)$} %% подпись к рисунку
			\label{fig:fig2_2}
		\end{minipage}
	\end{center}
\end{figure}

When implementing the algorithm for finding the attraction areas of the local  minima, we used the algorithms from OpenCV library to construct decision tree. OpenCV is an open source library of the algorithms of computer vision, image processing, and  general-purpose numerical algorithms. More details on the decision tree can be found  in \bblcite{Brahmbhatt2013}.

@@

%Ниже приведены результаты сравнения двух параллельных алгоритмов --- асинхронного GSA и его модификации с использованием decision tree и локального метода Хука-Дживса. Численное сравнение проводилось на классах функций Simple и Hard размерности 2, 3, 4 и 5 из \bblcite{Sergeyev2006}. Критерием остановки служило попадание точки очередного испытания trial в $\varepsilon$ окрестность истинного глобального минимума. 
%В таблице \ref{tab:2} представлено среднее количество итераций и ускорение по времени относительно последовательного запуска. Распараллеливание  осуществлялось с применением  MPI. При этом запуск производился на 4 процессах, а следовательно, на каждой итерации вычислялось 3 точки. Значение в скобках (если оно есть) показывает количество не решенных задач. Это означает, что было достигнут максимум по числу проводимых итераций, но в окрестность глобального минимума точка так и не попала.

Below, the results of comparing two parallel algorithms --- asynchronous global search algorithm  (AGSA) and its modification using the decision tree to find the attractor areas of the local minima  (Decision Tree) are presented. The numerical comparison was performed on two classes of the GKLS  functions (Simple and Hard from \bblcite{Sergeyev2006}) of dimensionalities 2, 3, 4, and 5. 
These two classes differ in the size of the attraction region of the global minimum point. 
For the simple class of problems, the radius of the attraction region is three times larger than for a complex one.

@@

In addition to using decision trees to identify attraction regions for local extrema of multiextremal functions, we also plan to use machine learning methods to separate the variables of the problem being solved. In many applied optimization problems the dependence of the objective function on some parameters is either linear or unimodal. Separating such variables into a special group and solving the problem using a parallel recursive optimization scheme \bblcite{Barkalov2020_1} can reduce the time for solving the problem by orders of magnitude compared to using a global search for all variables at once.

\vspace{1em}
{\it This study was supported by the Russian Science Foundation, project No.\,21-11-00204 .}

\insertbiblioen
